Hypertension and microbiome

### 2020-06-24 articletwo-sensitivityanalysis.Rmd
Final codes

@@

@@

>>>>

```{r load old, eval = FALSE}
load("session/session-sensitivity.Rdata")
```

@@

Self reported BP medication

```{r bp medication use}
pseq.species %>% meta %>% pull(BP_TREAT) %>% as.factor %>% summary %>% kable
```

@@

Excluding participants with diabtes, CAD, and any cancer

```{r only healthy}
pseq.species %>% meta %>%
    filter(PREVAL_DIAB == 1 | PREVAL_CHD == 1 | PREVAL_CR_ANYCANC == 1) %>%
    summarize(diab = sum(PREVAL_DIAB == 1),
              chd = sum(PREVAL_CHD == 1),
              canc = sum(PREVAL_CR_ANYCANC == 1),
              any = n()) %>%
    kable
```

### 2020-06-24 articletwo.Rmd
Final codes

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(egg)
library(gridExtra)
library(png)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(gtable)
library(car)
library(M3C)
library(emmeans)
library(DESeq2)
library(ggeffects)
library(pander)
library(hexbin)
library(readr)
```

@@

# Sources

@@

```{r data}
pseq.species <- import_filter_data("data/phfinrisk_species_all_drop50k_2018-12-21.RDs")
pseq.genus <- import_filter_data("data/phfinrisk_genus_all_drop50k_2018-11-16.RDs")
pseq.genus.coretaxa <- coretaxa(pseq.genus, detection = 0.1/100, prevalence = 1/100)
```

@@

### Bray curtis distance matrix

@@

### Beta diversity

@@

### PCoA

@@

### DeSeq2

```{r deseq2}
if (!file.exists("rds/dds.rds") || !file.exists("rds/dds3.rds") || !file.exists("rds/ddsNa.rds")) {
    pseq.genus.core.deseq <-  prune_taxa(pseq.genus.coretaxa, pseq.genus)

    if (!file.exists("rds/dds.rds")) {
        dds <- lapply(c2l(var.BP), function(x, pseq) {
            dds.data <- phyloseq_to_deseq2(pseq, deseq.formula(x, var.CL))
            DESeq(dds.data,
                  test="Wald",
                  fitType="parametric",
                  parallel = TRUE,
                  BPPARAM=MulticoreParam(16))
        }, pseq = pseq.genus.core.deseq)
        saveRDS(dds, file = "rds/dds.rds")
    }
    if (!file.exists("rds/dds3.rds")) {
        dds3 <- lapply(deseq.list(var.CL, var.CL.min), function(x, pseq) {
            dds.data <- phyloseq_to_deseq2(pseq, deseq.formula(x, "HYPERTENSION"))
            DESeq(dds.data,
                  test="Wald",
                  fitType="parametric",
                  parallel = TRUE,
                  BPPARAM=MulticoreParam(16))
        }, pseq = pseq.genus.core.deseq)
        saveRDS(dds3, file = "rds/dds3.rds")
    }
    if (!file.exists("rds/ddsNa.rds")) {
        ddsNa <- mclapply(c2l(var.BP), function(x) {
            myDESeq(pseq = pseq.species,
                    vars = c(x, var.CL),
                    coreterm = "^Lactobacillus.*Bacteria")
        }, mc.cores = length(var.BP))
        saveRDS(ddsNa, file = "rds/ddsNa.rds")        
    }
} else {
    dds <- readRDS(file = "rds/dds.rds")
    dds3 <- readRDS(file = "rds/dds3.rds")
    ddsNa <- readRDS(file = "rds/ddsNa.rds")
}
```

# Results

## Characteristics

@@

## Microbial diversity

@@

## Principal coordinate analysis

@@

## Associations for genera

@@

### Associations for hypertension looping covariates

@@

## Associations for Lactobacillus species

```{r lactobacillus species results}
dset.deseq.signf.na <- deseqresults(ddsNa, names.dset)
g.deseq.na <- deseqheatmap(dset.deseq.signf.na)
```

@@

## Salt subset

## Association between genera and dU-NA

```{r lacto deseq calculation}
pseq.genus.salt <- pseqsubset(pseq.genus, coretaxa = pseq.genus.coretaxa, saltsubset = TRUE)
```

Lactobacillus prevalence in subsample `r sprintf("%.1f%%", 100*pseq_prevalence(pseq.genus.salt))`. 
Levels of dU-Na `r sprintf("%.1f (%.1f)", mean(meta(pseq.genus)$dUNA, na.rm = TRUE), 
	sd(meta(pseq.genus)$dUNA, na.rm = TRUE))`

```{r significant results for salt subset}
dds.salt <- myDESeq(pseq = pseq.genus,
                    coretaxa = pseq.genus.coretaxa,
                    vars = c("dUNA", var.CL),
                    saltsubset = TRUE)

dds.salt %>%
    results(., name = "dUNA", tidy = TRUE) %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH"),
                  lfc_se = sprintf("%.4f±%.4f", log2FoldChange, lfcSE)) %>%
    select(row, lfc_se, qval) %>%
    filter(row == "Lactobacillus (Bacteria)") %>%
    kable
```

@@

```{r salt subset char}
df.genus.salt <- saltdset(pseq.genus.salt, dds.salt)
```

@@

## Association between Lactobacilli and dU-NA

```{r species salt}
dds.salt.species <- myDESeq(pseq = pseq.species,
                        vars = c("dUNA", var.CL),
                        coreterm = "^Lactobacillus.*Bacteria",
                        saltsubset = TRUE)
```

Significant results after FDR correction

```{r species salt results}
dds.salt.species %>%
    results(., name = "dUNA", tidy = TRUE) %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH"),
                  row = renametaxa(row)) %>%
    filter(qval < 0.05) %>%
    mutate(qval_exp = sprintf("%.4e", qval)) %>%
    kable
```

Prevalence for L. paracasei is
`r pseq_prevalence_format(pseq.species, "Lactobacillus_paracasei (Bacteria)")`.
and for L. salivarius is 
`r pseq_prevalence_format(pseq.species, "Lactobacillus_salivarius (Bacteria)")`.

```{r species plot}
df.species.salt <- deseq.abundances(dds.salt.species) %>%
    full_join(deseq.abundances(dds.salt), by = "sampleid") %>%
    left_join(meta(pseq.species) %>% tibble::rownames_to_column("sampleid"), by = "sampleid") %>%
    mutate_at(vars(contains('Lactobacillus')), .funs = list(group = ~factor(dplyr::ntile(., 4)))) %>%
    select(dUNA,
           Lactobacillus_paracasei.Bacteria_group,
           Lactobacillus_salivarius.Bacteria_group,
           Lactobacillus.Bacteria_group)

df.species.salt %>%
    saltboxplot %>%
    gather(taxa, abundance, -dUNA) %>%
    ggsave(file = "cache/boxplot-lacto-salt.png", plot = ., height = 3, width = 3.5, dpi = 300, unit = "in")
```

@@

Urinary sodium quartiles for L. paracei

```{r species salt levels paracasei, echo = FALSE}
df.species.salt %>%
    group_by(Lactobacillus_paracasei.Bacteria_group) %>%
    summarize(mean = round(mean(dUNA), 1),
              sd = round(sd(dUNA), 1)) %>%
    kable
```

Urinary sodium quartiles for L. salivarius

```{r salt levels salivarius, echo = FALSE}
df.species.salt %>%
    group_by(Lactobacillus_salivarius.Bacteria_group) %>%
    summarize(mean = round(mean(dUNA), 1),
              sd = round(sd(dUNA), 1)) %>%
    kable
```

>>>>

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(egg)
library(gridExtra)
library(png)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(gtable)
library(car)
library(M3C)
library(emmeans)
library(DESeq2)
library(ggeffects)
library(pander)
library(hexbin)
library(readr)
library(KEGGREST)
```

@@

# Import helper functions

@@

```{r data}
pseq.species <- import_filter_data("data/phfinrisk_species_all_drop50k_2018-12-21.RDs")
pseq.genus <- import_filter_data("data/phfinrisk_genus_all_drop50k_2018-11-16.RDs")
pseq.genus.coretaxa <- coretaxa(pseq.genus, detection = 0.1/100, prevalence = 1/100)
pseq.genus.core.deseq <-  prune_taxa(pseq.genus.coretaxa, pseq.genus)
```

@@

## Bray curtis distance matrix

@@

## Beta diversity

@@

## PCoA

@@

## DeSeq2

Associations for common genera

```{r deseq2}
if (!file.exists("rds/dds.rds")) {
    dds <- lapply(c2l(var.BP), function(x, pseq) {
        dds.data <- phyloseq_to_deseq2(pseq, deseq.formula(x, var.CL))
        DESeq(dds.data,
              test="Wald",
              fitType="parametric",
              parallel = TRUE,
              BPPARAM=MulticoreParam(16))
    }, pseq = pseq.genus.core.deseq)
    saveRDS(dds, file = "rds/dds.rds")
} else {
    dds <- readRDS(file = "rds/dds.rds")
}
```

The affect of third covariate for hypertension

```{r deseq2 third covariate}    
if (!file.exists("rds/dds3.rds")) {
    dds3 <- lapply(deseq.list(var.CL, var.CL.min), function(x, pseq) {
        dds.data <- phyloseq_to_deseq2(pseq, deseq.formula(x, "HYPERTENSION"))
        DESeq(dds.data,
              test="Wald",
              fitType="parametric",
              parallel = TRUE,
              BPPARAM=MulticoreParam(16))
    }, pseq = pseq.genus.core.deseq)
    saveRDS(dds3, file = "rds/dds3.rds")
}  else {
    dds3 <- readRDS(file = "rds/dds3.rds")
}
```

Associations for Lactobacili

```{r deseq2 lactobacilli}
if (!file.exists("rds/ddsLactobacilli.rds")) {
    ddsLactobacilli <- mclapply(c2l(var.BP), function(x) {
        myDESeq(pseq = pseq.species,
                vars = c(x, var.CL),
                coreterm = "^Lactobacillus.*Bacteria")
    }, mc.cores = length(var.BP))
    saveRDS(ddsLactobacilli, file = "rds/ddsLactobacilli.rds")
} else {
    ddsLactobacilli <- readRDS(file = "rds/ddsLactobacilli.rds")
}
```

Subgroup analyses for Lactobacilli

```{r heavy subgroup analyses for lactobasilli-BP}
if (!file.exists("rds/ddsSubLactobacilli.rds")) {
    ddsSubLactobacilli <- lapply(myDESeq.subsets(), function(subset) {
        mclapply(c2l(var.BP), function(bp) {
            myDESeq(pseq = pseq.species,
                    vars = bp %union% var.CL %difference% subset$exclude,
                    coreterm = "^Lactobacillus.*Bacteria",
                    FUN = subset$fun)
        }, mc.cores = length(var.BP))
    })
    saveRDS(ddsSubLactobacilli, file = "rds/ddsSubLactobacilli.rds")
} else {
    ddsSubLactobacilli <- readRDS(file = "rds/ddsSubLactobacilli.rds")
}

#ddsSubLactobacilli <-  readRDS(file = "rds/ddsSub-temp.rds") %>%
#    purrr::list_modify(drug = readRDS(file = "rds/ddsSubLactobacilli.rds")$drug)
```

# Characteristics

@@

<details><summary>Chi-square tests and t-tests</summary>

```{r chi squared}
lapply(c2l(var.CL, var.BP), function(x) {
    meta(pseq.species) %>%
        mutate(nagroup = ifelse(is.na(dUNA), 1, 0)) %$%
        { if (x %in% c("BL_AGE", "BMI", "MAP", "SYSTM", "DIASM", "PULSEPRESSURE"))
              a <<- t.test(as.formula(sprintf("%s ~ nagroup", x))) 
          else
              table(nagroup, .[[x]]) %>%
                  chisq.test() } %>%
        tidy}) %>%
    map_df(~.x, .id = "term") %>%
    mutate (p.value = pub.p(p.value)) %>%
    select(term, p.value)
```

```{r chi squared multilevel}
lapply(c2l(1, 2, 3), function(x) 
    meta(pseq.species) %>%
    mutate(nagroup = ifelse(is.na(dUNA), 1, 0),
           Q57X = ifelse(Q57X == x, 1, 0)) %$%
    table(nagroup, Q57X) %>%
    chisq.test() %>%
    tidy) %>%
    map_df(~.x, .id = "term") %>%
    mutate (p.value = pub.p(p.value)) %>%
    select(term, p.value)
```

# Alpha and beta diversity plots

@@

# Principal coordinate analysis

@@

# Associations between genera and blood pressure indices

@@

## N:o associations for hypertension per included 3rd covariate

@@

# Associations between Lactobacilli and blood pressure indices

```{r lactobacillus species results}
dset.deseq.signf.na <- deseqresults(ddsLactobacilli, names.dset)
g.deseq.na <- deseqheatmap(dset.deseq.signf.na)
```

@@

# Association between Lactobacillus genus and urinary sodium

```{r significant results for salt subset}
(dds.salt <- myDESeq(pseq = pseq.genus,
                    coretaxa = pseq.genus.coretaxa,
                    vars = c("dUNA", var.CL),
                    saltsubset = TRUE)) %>%
    results(., name = "dUNA", tidy = TRUE) %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH"),
                  lfc_se = sprintf("%.4f±%.4f", log2FoldChange, lfcSE)) %>%
    select(row, lfc_se, qval) %>%
    filter(row == "Lactobacillus (Bacteria)") %>%
    kable
```

@@

```{r salt subset char}
pseq.genus.salt <- pseqsubset(pseq.genus, coretaxa = pseq.genus.coretaxa, saltsubset = TRUE)
df.genus.salt <- saltdset(pseq.genus.salt, dds.salt)
```

@@

# Association between Lactobacilli and urinary sodium

Analysis between *Lactobacilli* and urinary salt.

```{r species salt}
( dds.salt.species <- myDESeq(pseq = pseq.species,
                        vars = c("dUNA", var.CL),
                        coreterm = "^Lactobacillus.*Bacteria",
                        saltsubset = TRUE) ) %>%
    results(., name = "dUNA", tidy = TRUE) %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH"),
                  row = renametaxa(row)) %>%
    filter(qval < 0.05) %>%
    mutate(qval_exp = sprintf("%.4e", qval)) %>%
    kable
```

```{r test}
dds.salt.species %>%
      results(., name = "dUNA", tidy = TRUE) %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH"),
                  row = renametaxa(row)) %>%
                                        #    filter(qval < 0.05) %>%
    arrange(qval) %>%
    mutate(qval_exp = sprintf("%.4e", qval)) %>%
    kable
```

```{r species salt}
dds.salt.species %>%
    results(., name = "SEX_1_vs_0", tidy = TRUE) %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH"),
                  row = renametaxa(row)) %>%
    filter(qval < 0.05) %>%
    mutate(qval_exp = sprintf("%.4e", qval)) %>%
    kable
```

The box plot for *Lactobacilli* groups and urinary salt to demonstrate the association.

```{r species plot}
( df.species.salt <- deseq.abundances(dds.salt.species) %>%
    full_join(deseq.abundances(dds.salt), by = "sampleid") %>%
    left_join(meta(pseq.species) %>% tibble::rownames_to_column("sampleid"), by = "sampleid") %>%
    mutate_at(vars(contains('Lactobacillus')), .funs = list(group = ~factor(dplyr::ntile(., 4)))) %>%
    select(dUNA,
           Lactobacillus_paracasei.Bacteria_group,
           Lactobacillus_salivarius.Bacteria_group,
           Lactobacillus.Bacteria_group) ) %>%
    gather(taxa, abundance, -dUNA) %>%
    saltboxplot %>%
    ggsave(file = "cache/boxplot-lacto-salt.png", plot = ., height = 3, width = 3.5, dpi = 300, unit = "in")
```

@@

Mean and SD for urinary sodium quartiles

```{r species salt levels paracasei, echo = FALSE}
lapply(c2l(df.species.salt %>% select(contains("Lactobacillus")) %>% colnames),
       function(x) df.species.salt %>%
                   dplyr::rename(group = !!x) %>%
                   group_by(group) %>%
                   summarize(mean = round(mean(dUNA), 2),
                             sd = round(sd(dUNA), 2)) ) %>%
    map_df(~as.data.frame(.x), .id = "taxa")  %>%
    mutate(taxa = renametaxa(taxa)) %>%
    kable
```

# Subgroup analyses

## Sex

```{r deseq dual plot, echo = FALSE}
g.ddssub.male <- ddsSubLactobacilli$male %>%
    deseqresults(., names.dset) %>%
    deseqheatmap(breaks = seq(-2, 2)) %>%
    { . + theme(legend.position = "none",
                plot.title = element_text(size = 12, face = "plain")) } %>%
    myggtitle(., title= "A. Men")

g.ddssub.female <- ddsSubLactobacilli$female %>%
    deseqresults(., names.dset) %>%
    deseqheatmap(breaks = seq(-2, 2)) %>%
    { g.ddssub.female.raw <<- . } %>%
    { . + theme(legend.position = "none",
                 plot.title = element_text(size = 12, face = "plain")) } %>%
    myggtitle(., title= "B. Women")

g.ddssub.legend <- get_legend(g.ddssub.female.raw) %>% as_ggplot

g.deseq.subsample.sex <- arrangeGrob(g.ddssub.male,
                                     g.ddssub.female,
                                     g.ddssub.legend ,
                                     ncol = 3,
                                     widths = c(2, 2, 1))
ggsave(file = "cache/deseq-subsample-sex.png",
       plot = g.deseq.subsample.sex,
       width = 17,
       height = 11,
       dpi = 300,
       unit = "cm")
```

## Antihypertensive

```{r deseq dual plot, echo = FALSE}
g.ddssub.nodrug <- ddsSubLactobacilli$nodrug %>%
    purrr::list_modify(HYPERTENSION = NULL) %>%
    deseqresults(., names.dset) %>%
    deseqheatmap(breaks = seq(-2, 2)) %>%
    { . + theme(legend.position = "none",
                 plot.title = element_text(size = 12, face = "plain")) } %>%
    myggtitle(., title= "A. No antihypertensive medication use")

g.ddssub.drug <- ddsSubLactobacilli$drug %>%
    purrr::list_modify(HYPERTENSION = NULL) %>%
    deseqresults(., names.dset) %>%
    deseqheatmap(breaks = seq(-2, 2)) %>%
    { . + theme(legend.position = "none",
                plot.title = element_text(size = 12, face = "plain")) } %>%
    myggtitle(., title= "B. Antihypertensive medication use")

g.deseq.subsample.drug <- arrangeGrob(g.ddssub.nodrug,
                                      g.ddssub.drug,
                                      g.ddssub.legend,
                                     ncol = 3,
                                     widths = c(2, 2, 1))
ggsave(file = "cache/deseq-subsample-drug.png",
       plot = g.deseq.subsample.drug,
       width = 19,
       height = 14.2,
       dpi = 300,
       unit = "cm")
```

# Sodium subsample

```{r subset salt sex}
( ddsSubLactobacilliSodium <- mclapply(myDESeq.subsets(), mc.cores = 4, function(subset)
    myDESeq(pseq = pseq.species,
            vars = "dUNA" %union% var.CL %difference% subset$exclude,
            coreterm = "^Lactobacillus.*Bacteria",
            saltsubset = TRUE,
            FUN = subset$fun)) ) %>%
    myDESeq.tidy() %>%
    map_df(~.x, .id = "subgroup") %>%
    kable
```

```{r subset salt sex}
ddsSubLactobacilliSodium %>%
    myDESeq.tidy()) %>%
    map_df(~.x, .id = "subgroup") %>%
    kable
```

## Lactobacilli prevalences

Prevalence

```{r prevalence 0.01%}
lapply(pseq.species %>% taxa %>% mygrep(word = "^Lactobacillus_", .) %>% c2l, function(x) 
    pseq_prevalence(pseq.species, taxa = x, limit = 0.01/100)) %>%
    tibble::enframe() %>% 
    unnest(cols = value) %>%
    arrange(desc(value)) %>%
    kable

females <- pseq.species %>%
    meta %>%
    tibble::rownames_to_column("sampleid") %>%
    filter(SEX == 1) %>%
    pull(sampleid)

males <- pseq.species %>%
    meta %>%
    tibble::rownames_to_column("sampleid") %>%
    filter(SEX == 0) %>%
    pull(sampleid)

taxa(pseq.species) %>% mygrep(word = "Salivarius")

pseq_prevalence(prune_samples(males, pseq.species), limit = 0.01/100, taxa = "Lactobacillus_paracasei (Bacteria)")
pseq_prevalence(prune_samples(females, pseq.species), limit = 0.01/100, taxa = "Lactobacillus_paracasei (Bacteria)")

```

```{r prevalence 0.1%}
lapply(c("Lactobacillus_paracasei (Bacteria)", "Lactobacillus_salivarius (Bacteria)") %>% c2l, function(x) 
    pseq_prevalence(pseq.species, taxa = x, limit = 0.1/100)) %>%
    tibble::enframe() %>% 
    unnest(cols = value) %>%
    arrange(desc(value)) %>%
    kable
```

# Functional analysis

```{r import KEGG data}
dset.ko  <- readRDS("data/function_activity_copied_20200213.rds")$ko %>%
    as.data.frame %>%
    rownames_to_column(var = "term") %>%
    gather(sampleid, value, -term) %>% 
    spread(term, value) %>%
    mutate_at(vars(starts_with("K")), ~scale(log10(. + 1))) %>%
    left_join(meta(pseq.species) %>% rownames_to_column(var = "sampleid"), by = "sampleid")
```

```{r model KEGG}
ret.ko <- loop.lm(dset = dset.ko,
               response = "SYSTM",
               loops = dset.ko %>% colnames %>% mygrep(word = "^K[0-9]+"),
               covariates = var.CL) %>%
    loop.results()
```

```{r process KEGG results}
ret.ko %>%
    filter(qval < 0.05) %>%
    arrange(qval) %>%
    mutate(mean_ci = sprintf("%.2f (%.2f to %.2f)", estimate, conf.low, conf.high),
           qval = pub.p(qval)) %>%
    mutate(desc = mykeggget(term)) %>%
    select(term, desc, mean_ci, qval) %>%
    kable
```

<details><summary>Diagnostic plots</summary>

```{r KEGG qq plots}
diagnosticqqplot(dset.ko, ret.ko %>% filter(qval < 0.05) %>% pull(term)) %>%
    ggsave(file = "cache/kegg.png", plot = ., height = 40, width = 10, dpi = 300, unit = "in")
```

<img src = "cache/kegg.png" />

```{r KEGG scatter plots}
diagnosticscatterplot(dset.ko,
                      ret.ko %>% filter(qval < 0.05) %>% pull(term),
                      SYSTM) %>%
    ggsave(file = "cache/kegg-scatter.png", plot = ., height = 40, width = 10, dpi = 300, unit = "in")
```

<img src = "cache/kegg-scatter.png" />

```{r process KEGG results}
ret.ko %>%
    filter(qval < 0.05, estimate < 0) %>%
    mutate(term = paste0("n-", term),
           size = paste0("v-", 1/qval)) %>%
    select(term, size) %>%
    write_tsv(path = "cache/functree_negative.tsv", col_names = FALSE)
```
